Automatic Code Optimizations on GPU Architectures

### 2018-01-09 Paper.tex
Abstract and Corrections

\begin{abstract}

\end{abstract}

@@

	Todays compilers for CPU architectures are optimized a lot. There are many generally known optimizations like loop unrolling and constant precalcuation. The aim is to avoid problems of CPUs, such as costly instructions like jump- and division-operations \cite{?}.

	Loop unrolling tries to get rid of jump-operations, that accelerates the assembler code by unrolling a loop to procedural code without jump-operations \cite{?}. Constant precalcuation tries to evaluate a formula as best as possible without known variables \cite{?}.

	Compilers for GPGPU applications, which are used to run general purpose applications on a GPU, have to solve different problems. One of the main problems is the memory access time, which increases, if multiple threads want to access the same part of memory. In this paper, we want to collect information about different optimizations for compilers to reduce time, that is spent on waiting for memory \cite{?}.

	It is hard to write performant code, which uses the GPU to accelerate computation. NVIDIA provides CUDA, which is a programming model for different programming languages, which makes it easier to utilize the GPU for certain application. Nevertheless, CUDA introduces six different memory types and multi-threading, which is not the daily business of a programmer. Furthermore, it is important and necessary to optimize code as best as possible. \cite{nvc}

@@

@@

	\subsubsection{Threads}
	Threads are subroutines, so a small set of instructions, which can be managed and executed independently from other threads \cite{?}. Threads typically have an independent stack for local variables and share heap with all other threads. Due to access from multiple independent threads, heap access must be managed.

	\subsubsection{Thread Blocks}
	Thread Blocks is a construct, which was introduced by CUDA to organize threads in groups. Threads, that are grouped in a thread block, can access a part of the storage simultaneously.

	\subsubsection{Hardware Implementation}
	NVIDIA GTX 1080 has 20 multiprocessors with 128 threads each. This means that the threads of 20 thread blocks can be executed simultaneously. If there are more than 20 thread blocks, they are executed one after the other \cite{1080gtxdepth}.
	That means, that 2560 threads can be executed in parallel.

@@

		% Intro with Point - I tried
		Big Computations needs a lot of calculation time. Those computations can be mostly splitted into multiple smaller workloads, which can be calculated seperately. \cite{GPGPUCompiler} While CPUs are designed to calculate bigger workloads with less threads, GPUs are designed to calculate smaller workloads on hundreds of threads. In its original form, each pixel could be calculated individually on a GPU, which implies heavy multi threading.\cite{myth}

@@

		\subsubsection{Threading} A problem usually contains multiple smaller workloads, that can be calculated independently on GPUs. This calculations can be processed parallel to save time. Modern GPUs have more than 2500 cores with up to 1.7 GHz, which can execute code parallel. Other papers determine speedups up to 20 times.\cite{1080gtx} \cite{speedup}

		\subsubsection{Data Transfer Time}  Transfering data from memory into GPU memory is usually a bottle lag. Calculations, which uses data from secondary storage or produces a lot of data, have to transfer data to the host system, that moves the data to CPU memory or on secondary storage. NVIDIA GTX 1080 has a data transfer rate of 10 Gbps, which is 1.25 GB/s. Based on this data it takes a 12,8 seconds to refill the entire 8 GB of storage on the graphics card. \cite{1080gtx} \cite{transferspeedup}

		\subsubsection{Complexity} Complexity is a huge difficulty. The greater the degree of complexity, the more difficult and slower development progresses. Often code is not optimized if the consumer can deal with the waiting time, because development of accelerated solutions would cost much money. \cite{?}

@@

		NVIDIA CUDA is a parallel computing platform and model, which was developed by NVIDIA to speed up applications by usage of NVIDIA GPUs. For example CUDA can be plugged as library into C programs to make use of a CUDA-capable GPU. \cite{cudacpp}

@@

		In Figure 1, we see a simple vector addition, that is accelerated by using GPU with CUDA. You have to define a kernel function, which is a simple c function, which is called $N$ times by $N$ different CUDA threads. This kernel function is executed on GPU. \texttt{threadIdx.x} contains, in this simple example, just the number (\texttt{ 0 $\leq$ threadIdx.x $<$ N}) of the thread, which called the function. In Figure 2, we see the mathematical equation, which is done by the code in Figure 1. We can calculate every $c_i$ independently in sepeate threads, which makes it possible to speed up the process. This example is trivial, however matrix operations, that are more complex, can be speed up with the same approach. \cite{cudap}
		\begin{figure}[htbp] 
			\begin{align*}
				\begin{pmatrix}
				c_{1} \\
				c_{2} \\
				\vdots \\
				c_{m}
				\end{pmatrix} 
				= 
				\begin{pmatrix}
				a_{1} \\
				a_{2} \\
				\vdots \\
				a_{m}
				\end{pmatrix}
				+ 
				\begin{pmatrix}
				b_{1} \\
				b_{2} \\
				\vdots \\
				b_{m}
				\end{pmatrix}			
			\end{align*}
			\caption{Mathematical Equation for Vector Addition}
		\end{figure}

@@

		\subsubsection{Register}
			Register Memory is quite similar to CPU Register Memory and is used as storage for local variables, etc. It is limited to 16kb and everything, which exceeds this limit, will be pushed on Local Memory. \cite{accsql}
		\subsubsection{Local Memory}
			Local Memory is an abstraction of CUDA, which makes it possible to store information in Global Memory, which is only available from current thread.\cite{accsql}
		\subsubsection{Shared Memory}
			Shared Memory is much faster than the Global Memory. It is visible to all threads, which are in the same thread block. It is intented to share data across multiple threads, that are in the same thread block.\cite{cudalite} \cite{cudap}
		\subsubsection{Global Memory}
			Global Memory is the slowest memory. It is visible to all threads, can be read and written from all threads and can be as General Purpose memory. It is very similar to CPU memory, which makes it easy to use to share data across multiple thread blocks. \cite{cudalite}
		\subsubsection{Constant Memory}
			Constant Memory is quite similar to the Global Memory, but can only be read. It can help to reduce overhead by caching information. \cite{cudalite} \cite{accsql}  
		\subsubsection{Texture Memory}
			Texture Memory is quite similar to the Global Memory, but can only be read. It provides some caching methods, which can be used to accelerate certain applications, but this is not used in the further explanations. \cite{cudap}

@@

	Fortunately, there are a few other techniques to accelerate GPGPU programs.

@@

	Thread Blocks have a shared memory segment, where they can save data for execution. If multiple thread blocks uses the same data from the global memory to execute its threads, we can merge them into one thread block to reduce the memory access for global memory.

@@

	In Figure 3, we see an example for merging two thread blocks. Both thread blocks contain a data segment, which is equivalent from global memory. After the merge the data segment is only fetched one time and the number of threads in the thread block adds up.

@@

	Thread merging is a technique where different threads are merged into one thread to reduce memory access times. In exchange, actual computing time gets lost.

@@

	With Thread merging, there is a possibility to merge different threads to reduce global memory access. 
	The goal of thread merging is to reuse data. In the best case, each block of data has to be fetched only once. In addition to thread block merging, which only allow reuse of Shared Data Segments, thread merging allows the reuse of shared registers too.

@@

	In Figure 4, we see an example for merging two thread blocks into one thread block by merging threads. Again, both blocks contain a shared data segment, which is equivalent. Additionally a shared register per thread is equivalent. By merging two threads into one, the shared data segment and the shared register just have to be fetched once.

@@

		Data prefetching optimizes the iteration over data structures by prefetching the data, that is needed in the following iteration step. In the current iteration step $i$ the data from iteration step $i+1$ gets fetched. This step allows you to overlap memory access time and computing time.

		In Figure 5, we see example code with data prefetching. To understand the concept the $tmp$ variable is important. In the beginning $tmp$ holds the first part of data. In the following iteration steps it requests the part of data, that is used in the following step. That is an asynchronous event, that run next to the calculations of the current iteration step.

@@

		In Figure 6, we see a timeline, that shows the overlapping execution times from fetching data and executing code after the optimization. There is no time saving in the first iteration, because it is not possible to execute code without loaded data. After the first iteration, data can be loaded and code can be executed at the same time.

@@

		The global memory of graphics cards is divided into partitions. Partitions can only be read or written by one thread at a time. If a record is loaded into global memory and is to be processed by multiple threads, then different parts of the record are processed by different threads.  If all parts of the data set are located on a few partitions, then many threads need to access a partition, so memory access is slower. This is called partition camping.

		If the data is evenly distributed across all available partitions, a single partition has fewer threads to access, which speeds up memory access.

@@

		In Figure 7, we see $\text{SM-1}, \text{SM-2}, \dots, \text{SM-30}$, which are different accessors.

		In the table titled "Without Partition Camping" we see that the data for the different accessors are distributed evenly over all partitions.

		In the table titled with "With Partition Camping" we see that the data is all in one partition and that all accessors would have to access one partition, resulting in long waiting times.

@@

@@

\begin{thebibliography}{00}
\bibitem{nvc} CUDA Zone - https://developer.nvidia.com/cuda-zone
\bibitem{GPGPUCompiler} Yi Yang, Ping Xiang, Jingfei Kong, Huiyang Zhou, "A GPGPU Compiler for Memory Optimization and Parallelism Management", 2010
\bibitem{myth} Victor W Lee, Changkyu Kim, Jatin Chhugani et. al "Debunking the 100X GPU vs. CPU Myth: An Evaluation of Throughput Computing on CPU and GPU", 2010
\bibitem{1080gtx} NVIDIA GeForce GTX 1080, https://www.nvidia.com/en-us/geforce/products/10series/geforce-gtx-1080/
\bibitem{1080gtxdepth} NVIDIA GeForce GTX 1080 White Paper,
https://international.download.nvidia.com/geforce-com/international/pdfs/GeForce\_GTX\_1080\_Whitepaper\_FINAL.pdf
\bibitem{cudacpp} GPU Accelerated Computing with C and C++ - https://developer.nvidia.com/how-to-cuda-c-cpp
\bibitem{cudap} Programming Model - Kernel http://docs.nvidia.com/cuda/cuda-c-programming-guide/index.html\#kernels
\bibitem{cudalite} Sain-Zee Ueng, Melvin Lathara, Sara S. Baghsorkhi, Wen-mei W. Hwu, "CUDA-Lite: Reducing GPU Programming", 2008
Complexity
\bibitem{accsql} Peter Bakkum, Kevin Skadron, "Accelerating SQL Database Operations on a GPU with CUDA", 2010
\bibitem{pc} Ashwin M. Aji, Mayank Daga, Wu-chun Feng, "Bounding the Effect of Partition Camping in GPU Kernels", 2011
\bibitem{mtransnvidia} Greg Ruetsch, Paulius Micikevicius, "Optimizing Matrix Transpose in CUDA", 2009

>>>>

\begin{abstract}
	GPU programming is a hot topic. More and more tasks requiring a lot of computing power are performed on graphics cards, such as bitcoin mining. Optimizations that make these applications even more efficient to take full advantage of the power of the graphics cards are all the more important.

	After a short introduction to parallel programming and NVIDIA CUDA, this paper compares different approaches to shortening the time a thread waits for memory. There are several optimizations, such as thread block merging, thread merging, data prefetching and the elimination of partition camping. These four optimizations are compared in four different categories.

	At the moment there are many implementations that offer improved compilers for CUDA, but few comparisons between different optimizations and their modes of action and lines of action. This paper aims to give an overview of the most important optimizations and to compare them in interesting and simple categories.
\end{abstract}

@@

	Todays compilers for CPU architectures are optimized a lot. There are many generally known optimizations like loop unrolling and constant precalcuation. The aim is to avoid problems of CPUs, such as costly instructions like jump- and division-operations \cite{gccoptimizations}.

	Loop unrolling tries to get rid of jump-operations, that in the best case accelerates the assembler code by unrolling a loop to procedural code without jump-operations \cite{gccoptimizations}. Constant precalcuation tries to evaluate a formula as best as possible without known variables \cite{gccoptimizations}.

	In general, these optimizations try to avoid cost-intensive operations. These are either avoided or calculated at compile time, so that the actual program runs faster.

	Compilers, which are used to create GPGPU applications, have to help the developer more than conventional CPU applications. The main problem with GPGPU applications is memory access. NVIDIA introduces CUDA, a programming model that allows developers to run code on graphics cards, but CUDA introduces six different types of memory, all of which have to be managed by the programmer, which is much more than traditional programming. In addition, this memory access is slow in case of incorrect handling \cite{nvc} \cite{cudap}.

	Obviously, other problems have to be solved than with CPU applications. In GPGPU applications, memory access to global memory is the main problem and a problem that can be solved in different ways.
	Therefore, we would like to discuss here what optimizations already exist for the GPU \cite{nvc} \cite{cudap}.

@@

	This section is intended to provide an introduction to the various terms used in this paper. The terms Thread and Thread Blocks and General Purpose Computation on GPUs are explained. In addition, NVIDIA CUDA and the different types of memory introduced by CUDA will be presented to help improve understanding of the topic.

@@

	\subsubsection{Threads}
	Threads are subroutines, so a small set of instructions, which can be managed and executed independently from other threads. Threads typically have an independent stack for local variables and share heap with all other threads. Due to access from multiple independent threads, heap access must be managed  \cite{threadDefinition}.

	\subsubsection{Thread Blocks}
	Thread Blocks is a construct, which was introduced by CUDA to organize threads in groups. Threads, that are grouped in a thread block, can access a part of the storage simultaneously \cite{cudap}.

	\subsubsection{Hardware Implementation}
	NVIDIA GTX 1080 has 20 multiprocessors with 128 threads each. This means that the threads of 20 thread blocks can be executed simultaneously. If there are more than 20 thread blocks, they are executed one after the other.
	That means, that 2560 threads can be executed in parallel \cite{1080gtxdepth}.

@@

		% Intro with Point - I tried
		Big Computations need a lot of calculation time. Those computations can be mostly splitted into multiple smaller workloads, which can be calculated independently \cite{GPGPUCompiler}. While CPUs are designed to calculate bigger workloads with less threads, GPUs are designed to calculate smaller workloads on hundreds of threads. In its original form, each pixel could be calculated individually on a GPU, which implies heavy multi threading \cite{myth}.

@@

		\subsubsection{Threading} A problem usually contains multiple smaller workloads, that can be calculated independently on GPUs. This calculations can be processed parallel to save time. Modern GPUs have more than 2500 cores with up to 1.7 GHz, which can execute code parallel. Other papers determine speedups up to 20 times \cite{1080gtx} \cite{speedup}.

		\subsubsection{Data Transfer Time}  Transferring data from memory into GPU memory is usually a bottleneck. Calculations, which use data from secondary storage or produce a lot of data, have to transfer data to the host system, that moves the data to CPU memory or on secondary storage.

		The problem here is the PCIe connection, which is currently specified with a maximum data rate of 32 GB per second. Thus it takes at least 250ms to fill the entire memory of a current graphics card, which is relatively slow in the context of fast calculations \cite{1080gtx} \cite{transferspeedup}.

		\subsubsection{Complexity} Complexity is a big difficulty. The greater the degree of complexity, the more difficult and slower the development process.  Especially when programming GPUs, the complexity is high and the quality of the programming is decisive for the speed of the application. However, code is often not optimized when the consumer can cope with the waiting time, because the development of accelerated solutions would cost a lot of money \cite{cudalite}.

@@

		NVIDIA CUDA is a parallel computing platform and model, which was developed by NVIDIA to speed up applications by usage of NVIDIA GPUs. For example CUDA can be plugged as library into C programs to make use of a CUDA-capable GPU \cite{cudacpp}.

@@

		In Figure 1, we show a simple vector addition, that is accelerated by using GPU with CUDA. You have to define a kernel function, which is a simple c function, which is called $N$ times by $N$ different CUDA threads. This kernel function is executed on GPU. \texttt{threadIdx.x} contains, in this simple example, just the number (\texttt{ 0 $\leq$ threadIdx.x $<$ N}) of the thread, which called the function. In Figure 2, we show the mathematical equation, which is done by the code in Figure 1. We can calculate every $c_i$ independently in sepeate threads, which makes it possible to speed up the process. This example is trivial, however matrix operations, that are more complex, can be speed up with the same approach \cite{cudap}.
		\begin{figure}[htbp] 
			\begin{align*}
				\begin{pmatrix}
				c_{1} \\
				c_{2} \\
				\vdots \\
				c_{m}
				\end{pmatrix} 
				= 
				\begin{pmatrix}
				a_{1} \\
				a_{2} \\
				\vdots \\
				a_{m}
				\end{pmatrix}
				+ 
				\begin{pmatrix}
				b_{1} \\
				b_{2} \\
				\vdots \\
				b_{m}
				\end{pmatrix}			
			\end{align*}
			\caption{Mathematical Equation for Vector Addition}
		\end{figure}

@@

		\subsubsection{Register}
			Register Memory is quite similar to CPU Register Memory and is used as storage for local variables, etc. It is limited to 16kb and everything, which exceeds this limit, will be pushed on Local Memory \cite{accsql}.
		\subsubsection{Local Memory}
			Local Memory is an abstraction of CUDA, which makes it possible to store information in Global Memory, which is only available from current thread \cite{accsql}.
		\subsubsection{Shared Memory}
			Shared Memory is much faster than the Global Memory. It is visible to all threads, which are in the same thread block. It is intented to share data across multiple threads, that are in the same thread block \cite{cudalite} \cite{cudap}.
		\subsubsection{Global Memory}
			Global Memory is the slowest memory. It is visible to all threads, can be read and written from all threads and can be as General Purpose memory. It is very similar to CPU memory, which makes it easy to use to share data across multiple thread blocks \cite{cudalite}.
		\subsubsection{Constant Memory}
			Constant Memory is quite similar to the Global Memory, but can only be read. It can help to reduce overhead by caching information \cite{cudalite} \cite{accsql}.
		\subsubsection{Texture Memory}
			Texture Memory is quite similar to the Global Memory, but can only be read. It provides some caching methods, which can be used to accelerate certain applications, but this is not used in the further explanations \cite{cudap}.

@@

	The primary objective is to keep the time for memory access as short as possible. There are various optimizations for this, such as thread block and thread merging, data prefetching and the elimination of partition camping.

@@

	Thread Blocks have a shared memory segment, where they can save data for execution. If multiple thread blocks use the same data from the global memory to execute its threads, we can merge them into one thread block to reduce the memory access for global memory \cite{GPGPUCompiler}.

@@

	In Figure 3, we show an example for merging two thread blocks. Both thread blocks contain a data segment, which is equivalent from global memory. After the merge the data segment is only fetched one time and the number of threads in the thread block adds up.

@@

	Thread merging is a technique where different threads are merged into one thread to reduce memory access times. In exchange, there are less concurrent threads, which reduces the absolute computation time.

@@

	The goal of thread merging is to reuse data. In the best case, each block of data has to be fetched only once. In addition to thread block merging, which only allows for the reuse of Shared Data Segments, thread merging allows the reuse of shared registers too \cite{GPGPUCompiler}.

@@

	In Figure 4, we show an example for merging two thread blocks into one thread block by merging threads. Again, both blocks contain a shared data segment, which is equivalent. Additionally a shared register per thread is equivalent. By merging two threads into one, the shared data segment and the shared register just have to be fetched once.

@@

		Data prefetching optimizes the iteration over data structures by prefetching the data, that is needed in the following iteration step. In the current iteration step $i$ the data from iteration step $i+1$ gets fetched. This step allows you to overlap memory access time and computing time \cite{GPGPUCompiler}. In exchange the actual iteration step become slower by adding some extra instructions for prefetching.

		In Figure 5, we show example code with data prefetching. The code is part of the code for matrix multiplication. To understand the concept the $tmp$ variable is important. In the beginning $tmp$ holds the first part of data. In the following iteration steps it requests the part of data, that is used in the following step. That is an asynchronous event, that run next to the calculations of the current iteration step.

@@

		In Figure 6, we show a timeline, that shows the overlapping execution times from fetching data and executing code after the optimization. There is no time saving in the first iteration, because it is not possible to execute code without loaded data. After the first iteration, data can be loaded and code can be executed at the same time.

@@

		The global memory of graphics cards is divided into partitions. Partitions can only be read or written by one thread at a time. If several threads want to access a partition at the same time, a queue is created and the threads cannot calculate any further. If all parts of the data set are located on a few partitions, then many threads need to access one partition, so memory access is slower. This is called partition camping \cite{pc} \cite{GPGPUCompiler}.

		If the data is evenly distributed across all available partitions, a single partition has fewer threads to access, which speeds up memory access. However, please note that data can only be efficiently redistributed if there is enough space and a good structure of the data.  In addition, it can happen that redistribution makes the code faster in one place, but decelerates other places where the data is used in a different way \cite{pc} \cite{GPGPUCompiler}.

@@

		In Figure 7, we show $\text{SM-1}, \text{SM-2}, \dots, \text{SM-30}$, which are different accessors.

		In the table titled "Without Partition Camping" we show that the data for the different accessors are distributed evenly over all partitions.

		In the table titled with "With Partition Camping" we show that the data is all in one partition and that all accessors would have to access one partition, resulting in long waiting times.

@@

	\begin{figure}[htbp]
		\centerline{
			\begin{tabular}{ c | c | c | c | c }
				Optimization & UC & E & D & EC \\ \hline
				Thread Block Merge & - & + & - & \\
				Thread Merge & - & + & - & \\
				Data Prefetching & + & - & + & \\
				Eliminiation of Partition Camping & + & o & + & \\
			\end{tabular}
		}
		\caption{Comparison of different optimizations - Use Cases (UC), Efficiency (E), Difficulty (D), Error Cases (EC)}
		\label{figtable}
	\end{figure}

@@

\begin{thebibliography}{00}
\bibitem{nvc} CUDA Zone - https://developer.nvidia.com/cuda-zone
\bibitem{GPGPUCompiler} Yi Yang, Ping Xiang, Jingfei Kong, Huiyang Zhou, "A GPGPU Compiler for Memory Optimization and Parallelism Management", 2010
\bibitem{myth} Victor W Lee, Changkyu Kim, Jatin Chhugani et. al "Debunking the 100X GPU vs. CPU Myth: An Evaluation of Throughput Computing on CPU and GPU", 2010
\bibitem{1080gtx} NVIDIA GeForce GTX 1080, https://www.nvidia.com/en-us/geforce/products/10series/geforce-gtx-1080/
\bibitem{1080gtxdepth} NVIDIA GeForce GTX 1080 White Paper,
https://international.download.nvidia.com/geforce-com/international/pdfs/GeForce\_GTX\_1080\_Whitepaper\_FINAL.pdf
\bibitem{cudacpp} GPU Accelerated Computing with C and C++ - https://developer.nvidia.com/how-to-cuda-c-cpp
\bibitem{cudap} Programming Model - Kernel http://docs.nvidia.com/cuda/cuda-c-programming-guide/index.html\#kernels
\bibitem{cudalite} Sain-Zee Ueng, Melvin Lathara, Sara S. Baghsorkhi, Wen-mei W. Hwu, "CUDA-Lite: Reducing GPU Programming", 2008
Complexity
\bibitem{accsql} Peter Bakkum, Kevin Skadron, "Accelerating SQL Database Operations on a GPU with CUDA", 2010
\bibitem{pc} Ashwin M. Aji, Mayank Daga, Wu-chun Feng, "Bounding the Effect of Partition Camping in GPU Kernels", 2011
\bibitem{mtransnvidia} Greg Ruetsch, Paulius Micikevicius, "Optimizing Matrix Transpose in CUDA", 2009 
\bibitem{gccoptimizations} GCC Optimize Options - https://gcc.gnu.org/onlinedocs/gcc/Optimize-Options.html
\bibitem{threadDefinition}  Lamport, Leslie (September 1979). "How to Make a Multiprocessor Computer That Correctly Executes Multiprocess Programs" (PDF). IEEE Transactions on Computers. C–28 (9): 690–691. doi:10.1109/tc.1979.1675439.
